---
title: "Simulating NFL seasons using nflseedR"
author: "Lee Sharpe"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ifelse(future::supportsMulticore(), future::plan("multicore"), future::plan("multisession"))
```

# Running a Simulation

Let's do a quick simulation! We'll apply the `fresh_season = TRUE` option,
which blanks out all of the game results and simulates everything from scratch.
This is a Monte Carlo style of simulation, meaning that there are seasons
simulated.

Loading the package is obligatory, so it is done first (along with `dplyr` for data wrangling and the `pipe`):

```{r setup}
library(nflseedR)
library(dplyr, warn.conflicts = FALSE)
```

By default, 1000 simulations are done. We'll set an initial random seed for
this demonstration so you can follow along and get the same results, but normally
you wouldn't do that.

```{r}
set.seed(4)
sims <- simulate_nfl(2020, fresh_season = TRUE)
```

The output contains a lot of pre-aggregated information, as well as the individual
results from each game of each simulation. For example, let's look at the overall
results for the Bears:

```{r}
sims$overall %>% dplyr::filter(team == "CHI") %>% knitr::kable()
```

We can see the Bears got 10.6 wins on average. They made the playoffs 82% of the time, won the division 35% of the time, and won the Super Bowl in 4%, and rarely received a top 5 draft pick. The `teams` section of the output will show how a team did in each simulated season. 

```{r}
sims$teams %>% dplyr::filter(team == "CHI") %>% utils::head(20) %>% knitr::kable()
```

Wow that second sim is wild! The Bears went 8-8, but then went on to win the Super Bowl
as the 7th seed! Let's check out those playoff games:

```{r}
sims$games %>% dplyr::filter(sim == 2, game_type != "REG") %>% knitr::kable()
```

In this simulation, as the 7th seed, the Bears played three road playoff games, defeating the Washington football team by 6 points in the wildcard round, the Vikings by 9 in the divisional around, and the Cowboys also by 9 in the NFC Conference Championship. Finally, they defeated the Chargers by 10 in the Super Bowl.

As you may have gathered at this point, the default simulation code picks a random
Elo for every team, and uses those as the starting Elo ratings for all 32 teams.
However, the default code Elo will adjust independently within each simulation as
each week is simulated. (The Elo model used is loosely based off of that of
[FiveThirtyEight](https://fivethirtyeight.com/methodology/how-our-nfl-predictions-work/).)

## Use Your Own Model

But of course the real value is putting in your own model into the simulator. To
accomplish this, you can write your own function which will determine the output of
games instead. As an example, here's a very stupid model that makes the team earlier
alphabetically win by 3 points 90% of the time, and lose by 3 points the other 10% of
the time.

```{r}
stupid_games_model <- function(t, g, w, ...) {
  # make the earlier alphabetical team win 90% of the time
  g <- g %>%
    dplyr::mutate(
      result = dplyr::case_when(
        week != w ~ result,
        away_team < home_team ~ sample(c(-3, 3), n(), prob = c(0.9, 0.1), replace = TRUE),
        away_team > home_team ~ sample(c(-3, 3), n(), prob = c(0.1, 0.9), replace = TRUE),
        TRUE ~ 0
      )
    )
  
  # return values
  return(list(teams = t, games = g))
}
```

When you create this function, the first two inputs are data on the teams (one row
per team per sim), and data on the games (one row per game per sim). The third argument
is the week number currently being simulated, as only one week is processed at a time.

Your function's job - by whatever means you choose - is to update the `result` column for
that week's games in each of the sims with the number of points the home team won by
(or lost by if negative, or 0 if the game ended in a tie).

It returns both the `teams` and the `games` data. It does both because this way you can
store information in new columns by team or by game to use in the next call. Make sure
your code both accepts and returns the appropriate information, or the simulator will
break!

For example, the default function updates a team's Elo after the game, and stores it
in the `teams` data. When the simulator processes the next week, it uses the updated Elo
rating to inform the team's next game.

!! Also, make sure you aren't overriding completed games or games that aren't in the current
week of `w`. The simulator will **not** stop you from setting past, present, or future
game results in your function, whether you meant to do so or not. !!

Let's run a simulation with `stupid_games_model` and see what happens:

```{r}
sims2 <- simulate_nfl(2020, process_games = stupid_games_model, fresh_season = TRUE)

sims2$overall %>% dplyr::arrange(team) %>% utils::head() %>% knitr::kable()
sims2$overall %>% dplyr::arrange(team) %>% utils::tail() %>% knitr::kable()
```

As you might expect, the earliest alphabetical teams win a lot. The Cardinals
won the Super Bowl in 68% of seasons! Meanwhile, the teams at the bottom
alphabetically are virtually certain to be at the top of the draft order.

## Adding In Your Own Data

This is all well and good, you might be thinking, but your model works off of
other data not in the simulator! How can that work? This is where we utilize R's
ability to have generic arguments.

The `...` at the end of the function definition means that the function can be
called with any number of additional arguments. You can name these whatever you
want, as long as they're not already the name of other defined arguments.

When you call the `simulate_nfl()` function, it too uses the `...` syntax, which
allows you to pass in any number of additional arguments to the function. The
simulator will in turn pass these on to *your* function that processes games.

For example, let's slightly modify our last example:

```{r}
biased_games_model <- function(t, g, w, ...) {
  
  # arguments
  args <- list(...)
  best <- ""
  worst <- ""
  
  # best team?
  if ("best" %in% names(args)) {
    best <- args$best
  }
  
  # worst team?
  if ("worst" %in% names(args)) {
    worst <- args$worst
  }

  # make the best team always win and the worst team always lose
  # otherwise, make the earlier alphabetical team win 90% of the time
  g <- g %>%
    dplyr::mutate(
      result = dplyr::case_when(
        week != w ~ result,
        away_team == best | home_team == worst ~ -3,
        away_team == worst | home_team == best ~ 3,
        away_team < home_team ~ sample(c(-3, 3), n(), prob = c(0.9, 0.1), replace = TRUE),
        away_team > home_team ~ sample(c(-3, 3), n(), prob = c(0.1, 0.9), replace = TRUE),
        TRUE ~ 0
      )
    )
  
  # return values
  return(list(teams = t, games = g))
}
```

This allows us to define `best` and `worst`, and use that information to
determine a result (in this case, have the best team always win and the 
worst team always lose). While `best` and `worst` are in this example
single-length character vectors, they can be data frames or any other R
data type.

Let's simulate using this:

```{r}
sims3 <- simulate_nfl(
  process_games = biased_games_model, 
  fresh_season = TRUE, 
  best = "CHI", 
  worst = "GB"
)
sims3$overall %>% dplyr::filter(team == "CHI") %>% knitr::kable()
sims3$overall %>% dplyr::filter(team == "GB") %>% knitr::kable()
```

And this shows exactly what we expect. By defining the Bears as the best team,
they always go 16-0, win the division, and win the Super Bowl. Interestingly,
they do not always get the #1 seed. This makes sense, however, as in games 
without the Bears or the Packers, the alphabetically earlier teams still wins 
90% of the time. The Cardinals would therefore be expected to go 16-0 in some
of the simulations, and in some of those have thee tiebreakers over the Bears
However, even in these simulations, they'll still lose to Bears in the end
when they meet in the playoffs.

Similarly, the Packers always go 0-16, and never have any kind of realistic
postseason. They do almost always get the #1 draft pick -- but not always. Using the same
logic as above, sometimes the Washington Football Team will go 0-16 too, and may
beat the Packers out for the #1 pick through tiebreakers.

## Simulation Configuration

There is a lot of flexibility in how you choose to run the simulation. These are
the parameters and how to configure them when you run the `simulate_nfl()` function.

* **nfl_season** - Which NFL season are you simulating? By default, it simulates
the most recent season for which the regular season schedule is available through
[Lee Sharpe's NFL game data](https://github.com/leesharpe/nfldata/tree/master/data).
This data only goes back to the 1999 season.
* **process_games** - This is where you supply a function you've written to encompass
your model used to determine simulated games results, like the examples above. By
default, this will generate a random Elo for every team per round of simulations, then
use that to determine game data.
* **playoff_seeds** - How many playoff seeds per conference are used? By default, this
is 7 for seasons 2020 and later, and 6 for earlier seasons.
* **if_ended_today** - This should only be used when running in the middle of the
regular season. It will take all completed games as done, but remove the rest of the
regular season games from the schedule, and begin the playoffs as if everything was
locked in based on the regular season data that exists so far.
* **fresh_season** - You'll see this was set to `TRUE` in all of our examples above.
This setting deletes any playoff games and clears out the results for all regular 
season games, so everything is generated fresh. The default is `FALSE` where all games
that have been completed in real life are treated as locked in, and instead remaining
games are simulated.
* **fresh_playoffs** - Similar to `fresh_season`, except instead when set to `TRUE`,
regular season results remain and only playoff games are deleted and then simulated.
The default is `FALSE` in which case playoff games that are completed are accepted
as they occurred,
* **tiebreaker_depth** - How far do you want tiebreakers to be utilized? Usually 
leaving it at the default below (`3`) is fine, but before the season starts, you
may wish to have less tie-breaking in order to 
  * `1`: All teams with the same record have any ties broken randomly.
  * `2`: Instead of evaluating common games if that step is reached, break any ties
randomly. But all earlier tiebreakers are handled correctly.
  * `3`: The default. All tiebreakers are handled through strength of schedule are
processed (or through strength of victory for draft pick order). In the unlikely
event of a further tie, it will be broken randomly.
* **simulations** - How many simulations should be run? Defaults to 1000.
* **sims_per_round** - The simulator can break things up into chunks of simulated
seasons, process each chunk on its own (called a round), and then aggregate everything
together at the end. The default value determines the number of locally available 
cores and calculates the number of simulations per round to be equal to half of 
the available cores (various benchmarks showed this results in optimal performance
in parallel processes). If your computer is hanging and forces a restart while 
running this simulation, it is recommended that you lower this number.
