---
title: "Get started with nflseedR"
author: "Sebastian Carl and Lee Sharpe"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Preface

nflseedR is designed to efficiently take over the sophisticated and complex rule set of the NFL regarding division ranks, postseason seeding and draft order. It is intended to be used for NFL season simulations to help modelers focus on their models rather than the tie-breaking procedures. The NFL's official procedures for breaking ties for postseason playoffs can be found [here](https://operations.nfl.com/the-rules/nfl-tie-breaking-procedures/) and [this site](https://operations.nfl.com/journey-to-the-nfl/the-nfl-draft/the-rules-of-the-draft/) explains the assigning of draft picks. 

However, it must be mentioned that nflseedR does not support all levels of tie-breakers at the moment. The deepest tie-breaker possible at the moment is the strength of schedule. After that, the decision is made at random. It should be noted, however, that the need for additional levels is extremely unlikely in reality.

# Using In-Simulation Functions

You can get NFL game data from this function:

* `load_sharpe_games()` to collect game information and results

And if you prefer, you can take or generate any set of game outcomes and let nflseedR
handle all of the NFL seeding and tiebreaker math for you with three in-simulation
functions (each can handle thousands of seasons at once):

* `compute_division_ranks()` to find all division winners,
* `compute_conference_seeds()` to get the playoff participants, and
* `compute_draft_order()` to find the resulting draft pick for each team.

The following sections will demonstrate how to use them and what input is required.

Loading the package is obligatory, so it is done first (along with `dplyr` for data wrangling and the `pipe`):

```{r setup}
library(nflseedR)
library(dplyr, warn.conflicts = FALSE)
options(digits = 3)
options(warn = -1)
```

## Load Sharpe Games

```{r}
games <- nflseedR::load_sharpe_games()
games %>% dplyr::slice_tail(n = 20) %>% knitr::kable()
```

This pulls game information from the games.rds file (equivalent to the games.csv file) from
[Lee Sharpe's NFL Data Github](https://github.com/nflverse/nfldata/tree/master/data)

## Find Division Ranks

This functions computes division ranks based on a data frame containing game results of one or more NFL seasons. So let's load some game data first (this example uses the game data of the 2012 and 2019 seasons):

```{r}
games <- nflseedR::load_sharpe_games() %>%
  dplyr::filter(season %in% c(2012, 2019)) %>%
  dplyr::select(sim = season, game_type, week, away_team, home_team, result)

dplyr::glimpse(games)
```

Please note the required column names:

* `sim` : A simulation ID. Normally 1 - n simulated seasons or (like in this case) just the year.

* `game_type` : One of 'REG', 'WC', 'DIV', 'CON', 'SB' indicating if a game was a regular season game or one of the playoff rounds.

* `week` : The week of the corresponding NFL season.

* `away_team` : Team abbreviation of the away team.

* `home_team` : Team abbreviation of the home team.

* `result` : Equals home score - away score.

Now the games data frame can be used to compute the division ranks (the parameter `.debug` is set to `TRUE` to show what the function is doing). 

```{r}
div_standings <- nflseedR::compute_division_ranks(games, .debug = TRUE)
dplyr::glimpse(div_standings)
```

Please note that the function outputs a list of data frames, the actual division standings as well as a data frame named `h2h`. The latter is an important input in the other functions (as it is used to break head-to-head ties) and can only be computed with `compute_division_ranks()`.

So here is the resulting division standings data frame for the 2012 season

```{r}
div_standings %>% 
  purrr::pluck("standings") %>% 
  dplyr::filter(sim == 2012) %>% 
  dplyr::select(division:div_rank) %>% 
  knitr::kable()
```

In that season the seconds division rank of the NFC South required a three way tie-breaker between the Panthers, Saints and Bucs. It was broken with the three-way Conference Record. This can be seen in the above given console output: `...DIV (3): Common Record` for the division rank number 2. The Bucs lost this tie-breaker with a 0.333 win percentage in the conference and the tie-breaking procedure goes on with a 2-way head-to-head comparison.

## Find Conference Seedings

This functions computes conference seedings based on the above computed division standings data frame. For efficiency reasons the above computed `h2h` data frame has to be passed to the function. The easiest way is to pass the list of data frames that is computed in the first step so we can do this (please note the number of playoff seeds):

```{r}
seeds <- div_standings %>% 
  nflseedR::compute_conference_seeds(h2h = .$h2h, playoff_seeds = 6, .debug = TRUE)
dplyr::glimpse(seeds)
```

Just like `compute_division_ranks()`, this function returns a list of two data frames so we can use it within a pipe. The resulting seeds for the 2012 season are given below.

```{r}
seeds %>% 
  purrr::pluck("standings") %>% 
  dplyr::filter(sim == 2012) %>% 
  dplyr::select(division:seed) %>% 
  knitr::kable()
```

## Find Draft Order

This function computes the draft order based on the playoff outcome and the regular season games. It requires all playoff results in the `games` data frame and the `game_type` of the Super Bowl has to be `"SB"`. For efficiency reasons the above computed `h2h` data frame has to be passed to the function as well. The easiest way is to pass the list of data frames that is computed in the above steps:

```{r}
draft <- seeds %>% 
  nflseedR::compute_draft_order(games = games, h2h = .$h2h, .debug = TRUE)
dplyr::glimpse(draft)
```

As this is the final step, the function `compute_draft_order` does not output `h2h` again. Instead it directly outputs the final standings including the draft order and the variable `exit` which indicates the week number of each team's final game (the Super Bowl Winner's `exit` equals 22):

```{r}
draft %>% 
  dplyr::filter(sim == 2012) %>% 
  dplyr::select(division:draft_order) %>% 
  knitr::kable()
```
