---
title: "Simulating NFL seasons using nflseedR"
author: "Lee Sharpe"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# ifelse(future::supportsMulticore(), future::plan("multicore"), future::plan("multisession"))
```

# Running a Simulation

Let's do a quick simulation! We'll apply the `fresh_season = TRUE` option,
which blanks out all of the game results and simulates everything from scratch.
This is a Monte Carlo style of simulation, meaning that there are seasons
simulated.

Loading the package is obligatory, so it is done first (along with `dplyr` for data wrangling and the `pipe`):

```{r setup}
library(nflseedR)
library(dplyr, warn.conflicts = FALSE)
options(digits = 3)
```

By default, 1000 simulations are done. We'll set an initial random seed for
this demonstration so you can follow along and get the same results, but normally
you wouldn't do that (for the same reason we set `sims_per_round = 1000`).

```{r}
set.seed(4)
sims <- simulate_nfl(2020, fresh_season = TRUE, sims_per_round = 1000)
```

The output contains a lot of pre-aggregated information, as well as the individual
results from each game of each simulation. For example, let's look at the overall
results for the Bears:

```{r}
sims$overall %>% dplyr::filter(team == "CHI") %>% knitr::kable()
```

We can see the Bears got 10.7 wins on average. They made the playoffs 82% of the time, won the division 35% of the time, and won the Super Bowl in 4%, and rarely received a top 5 draft pick. The `teams` section of the output will show how a team did in each simulated season. 

```{r}
sims$teams %>% dplyr::filter(team == "CHI") %>% utils::head(6) %>% dplyr::select(sim, team, wins, seed, draft_order) %>% knitr::kable()
```

Wow that second sim is wild! The Bears went 8-8, but then went on to win the Super Bowl
as the 7th seed! Let's check out those playoff games:

```{r}
sims$games %>% dplyr::filter(sim == 2, game_type != "REG") %>% knitr::kable()
```

In this simulation, as the 7th seed, the Bears played three road playoff games, defeating the Washington football team by 6 points in the wildcard round, the Vikings by 9 in the divisional around, and the Cowboys also by 9 in the NFC Conference Championship. Finally, they defeated the Chargers by 10 in the Super Bowl.

As you may have gathered at this point, the default simulation code picks a random
Elo for every team, and uses those as the starting Elo ratings for all 32 teams.
However, the default code Elo will adjust independently within each simulation as
each week is simulated. (The Elo model used is loosely based off of that of
[FiveThirtyEight](https://fivethirtyeight.com/methodology/how-our-nfl-predictions-work/).)

## Use Your Own Model

But of course the real value is putting in your own model into the simulator. To
accomplish this, you can write your own function which will determine the output of
games instead. As an example, here's a very stupid model that makes the team earlier
alphabetically win by 3 points 90% of the time, and lose by 3 points the other 10% of
the time.

```{r}
stupid_games_model <- function(teams, games, week_num, ...) {
  # make the earlier alphabetical team win 90% of the time
  games <- games %>%
    dplyr::mutate(
      result = dplyr::case_when(
        !is.na(result) | week != week_num ~ result,
        away_team < home_team ~ sample(c(-3, 3), n(), prob = c(0.9, 0.1), replace = TRUE),
        away_team > home_team ~ sample(c(-3, 3), n(), prob = c(0.1, 0.9), replace = TRUE),
        TRUE ~ 0
      )
    )
  
  # return values
  return(list(teams = teams, games = games))
}
```

When you create this function, the first two inputs are data on the teams (one row
per team per sim), and data on the games (one row per game per sim). The third argument
is the week number currently being simulated, as only one week is processed at a time.

Your function's job - by whatever means you choose - is to update the `result` column for
that week's games in each of the sims with the number of points the home team won by
(or lost by if negative, or 0 if the game ended in a tie).

It returns both the `teams` and the `games` data. It does both because this way you can
store information in new columns by team or by game to use in the next call. Make sure
your code both accepts and returns the appropriate information, or the simulator will
break!

For example, the default function updates a team's Elo after the game, and stores it
in the `teams` data. When the simulator processes the next week, it uses the updated Elo
rating to inform the team's next game.

!! Also, make sure you aren't overriding completed games or games that aren't in the current
week of `w`. The simulator will **not** stop you from setting past, present, or future
game results in your function, whether you meant to do so or not. !!

Let's run a simulation with `stupid_games_model` and see what happens:

```{r}
sims2 <- simulate_nfl(2020, process_games = stupid_games_model, fresh_season = TRUE, sims_per_round = 1000)

sims2$overall %>% dplyr::arrange(team) %>% utils::head() %>% knitr::kable()
sims2$overall %>% dplyr::arrange(team) %>% utils::tail() %>% knitr::kable()
```

As you might expect, the earliest alphabetical teams win a lot. The Cardinals
won the Super Bowl in 68% of seasons! Meanwhile, the teams at the bottom
alphabetically are virtually certain to be at the top of the draft order.

## Adding In Your Own Data

This is all well and good, you might be thinking, but your model works off of
other data not in the simulator! How can that work? This is where we utilize R's
ability to have generic arguments.

The `...` at the end of the function definition means that the function can be
called with any number of additional arguments. You can name these whatever you
want, as long as they're not already the name of other defined arguments.

When you call the `simulate_nfl()` function, it too uses the `...` syntax, which
allows you to pass in any number of additional arguments to the function. The
simulator will in turn pass these on to *your* function that processes games.

For example, let's slightly modify our last example:

```{r}
biased_games_model <- function(teams, games, week_num, ...) {
  
  # arguments
  args <- list(...)
  best <- ""
  worst <- ""
  
  # best team?
  if ("best" %in% names(args)) {
    best <- args$best
  }
  
  # worst team?
  if ("worst" %in% names(args)) {
    worst <- args$worst
  }

  # make the best team always win and the worst team always lose
  # otherwise, make the earlier alphabetical team win 90% of the time
  games <- games %>%
    dplyr::mutate(
      result = dplyr::case_when(
        !is.na(result) | week != week_num ~ result,
        away_team == best | home_team == worst ~ -3,
        away_team == worst | home_team == best ~ 3,
        away_team < home_team ~ sample(c(-3, 3), n(), prob = c(0.9, 0.1), replace = TRUE),
        away_team > home_team ~ sample(c(-3, 3), n(), prob = c(0.1, 0.9), replace = TRUE),
        TRUE ~ 0
      )
    )
  
  # return values
  return(list(teams = teams, games = games))
}
```

This allows us to define `best` and `worst`, and use that information to
determine a result (in this case, have the best team always win and the 
worst team always lose). While `best` and `worst` are in this example
single-length character vectors, they can be data frames or any other R
data type.

Let's simulate using this:

```{r}
sims3 <- simulate_nfl(
  nfl_season = 2020,
  process_games = biased_games_model, 
  fresh_season = TRUE, 
  best = "CHI", 
  worst = "GB",
  sims_per_round = 1000
)
sims3$overall %>% dplyr::filter(team == "CHI") %>% knitr::kable()
sims3$overall %>% dplyr::filter(team == "GB") %>% knitr::kable()
```

And this shows exactly what we expect. By defining the Bears as the best team,
they always go 16-0, win the division, and win the Super Bowl. Interestingly,
they do not always get the #1 seed. This makes sense, however, as in games 
without the Bears or the Packers, the alphabetically earlier teams still wins 
90% of the time. The Cardinals would therefore be expected to go 16-0 in some
of the simulations, and in some of those have thee tiebreakers over the Bears
However, even in these simulations, they'll still lose to Bears in the end
when they meet in the playoffs.

Similarly, the Packers always go 0-16, and never have any kind of realistic
postseason. They do almost always get the #1 draft pick -- but not always. Using the same
logic as above, sometimes the Washington Football Team will go 0-16 too, and may
beat the Packers out for the #1 pick through tiebreakers.

## Passing Data in from One Week to the Next

Sometimes though you want your data to keep updating as the simulation progresses.
For example, an Elo-based model that updates each team's Elo after each game. You
can pass in the starting Elo values per team, and as games are simulated, update
the Elo values for each team and store them in the `teams` data. This column will
be part of the `teams` data passed into your function when the following week is
simulated and your function is called.

Read the comments in the code below for specific tips on doing this but here are good ones:

* You can add columns to `teams` and/or `games` if you want.
* When doing joins to do the above, do left joins to make sure no rows are removed.
* Remove any "helper" columns you generate along the way you don't actually need before returning.
* Make sure any column doesn't get blindly joined on so it has `.x` and `.y` versions in Week 2 and R throws an error because an expected column name doesn't exist.
* Make sure you only update games with `is.na(result) & week == week_num`! You don't want to override completed games, or games from a week other than the current week being simulated.

```{r}
elo_model <- function(teams, games, week_num, ...) {

  # round out (away from zero)
  # this way the simulator never simulates a tie
  # the simulator will still allow ties to be simulated if you want
  # ... but not on playoff games
  round_out <- function(x) {
    x[!is.na(x) & x < 0] <- floor(x[!is.na(x) & x < 0])
    x[!is.na(x) & x > 0] <- ceiling(x[!is.na(x) & x > 0])
    return(x)
  }

  # we're going to store elo as a new columns in the teams data
  # it won't start off there of course, so we need to determine it
  # from our arguments
  if (!("elo" %in% colnames(teams))) {
    args <- list(...)
    if ("elo" %in% names(args)) {
      # pull the elo info from custom arguments
      teams <- teams %>%
        dplyr::inner_join(args$elo %>% dplyr::select(team, elo), by = c("team" = "team"))
    } else {
      # error with a friendly error message if no elo data is passed in
      stop("Pass in a tibble `elo` as an argument to `simulate_nfl()`")
    }
  }

  # isolate the ratings data by sim and by team only
  # we will want to join to the games data later and don't want excess columns
  ratings <- teams %>% dplyr::select(sim, team, elo)

  # simulate game outcomes
  games <- games %>%
    # add in the away team's elo to the game data
    # note we join on both `sim` and the team
    # always join on `sim` to make sure each sim cares about only its data
    dplyr::inner_join(ratings, by = c("sim" = "sim", "away_team" = "team")) %>%
    dplyr::rename(away_elo = elo) %>%
    # repeat for the home team as well
    dplyr::inner_join(ratings, by = c("sim" = "sim", "home_team" = "team")) %>%
    dplyr::rename(home_elo = elo) %>%
    dplyr::mutate(
      # calculate the elo difference
      elo_diff = home_elo - away_elo,
      # add in a small HFA amount if played at home
      elo_diff = elo_diff + ifelse(location == "Home", 20, 0),
      # make an adjustment for rest
      elo_diff = elo_diff + (home_rest - away_rest) / 7 * 25,
      # playoff games swing elo more
      elo_diff = elo_diff * ifelse(game_type == "REG", 1, 1.2),
      # from elo, we calculate the home team's win percentage
      wp = 1 / (10^(-elo_diff / 400) + 1),
      # we also can calculate the estimate (mean points home team wins by)
      estimate = elo_diff / 25,
      result = dplyr::case_when(
        # !!! ALWAYS DO THIS NEXT LINE IN YOUR `result` CHANGES !!!
        # you have to make sure you're only changing unfinished games in current week
        # if you don't do this, it will usually error out on a friendly error message
        is.na(result) & week == week_num ~ 
          as.integer(round_out(rnorm(n(), estimate, 13))),
        # if not this week or known result, leave as-is
        TRUE ~ as.integer(result)
      ),
      # simplify to 1 = win, 0 = loss, 0.5 = tie to help calculate elo shift
      outcome = dplyr::case_when(
        is.na(result) ~ NA_real_,
        result > 0 ~ 1,
        result < 0 ~ 0,
        TRUE ~ 0.5
      ),
      # calculate the amount to adjust home team's elo by
      elo_input = dplyr::case_when(
        is.na(result) ~ NA_real_,
        result > 0 ~ elo_diff * 0.001 + 2.2,
        result < 0 ~ -elo_diff * 0.001 + 2.2,
        TRUE ~ 1.0,
      ),
      elo_mult = log(pmax(abs(result), 1) + 1.0) * 2.2 / elo_input,
      elo_shift = 20 * elo_mult * (outcome - wp)
    ) %>%
    # we don't want these columns in `games` any more
    # remove any columns you don't need when you're done
    # otherwise the next week they'll get joined as `col.x` and `col.y`
    # which will almost certainly break your script
    dplyr::select(
      -away_elo, -home_elo, -elo_diff, -wp, -estimate,
      -outcome, -elo_input, -elo_mult
    )

  # apply elo shifts
  teams <- teams %>%
    # join games results from this week to away teams (within same sim!)
    # note this is a LEFT join, we don't want to remove any teams rows
    dplyr::left_join(games %>%
        dplyr::filter(week == week_num) %>%
        dplyr::select(sim, away_team, elo_shift),
      by = c("sim" = "sim", "team" = "away_team")
    ) %>%
    # away team's elo gets subtracted by elo amount
    # if the team wasn't an away team, do nothing
    dplyr::mutate(elo = elo - ifelse(!is.na(elo_shift), elo_shift, 0)) %>%
    # we don't want to keep `elo_shift` in `teams` either, remove it
    dplyr::select(-elo_shift) %>%
    # repeat the above except now do it for the home team
    dplyr::left_join(games %>%
        dplyr::filter(week == week_num) %>%
        dplyr::select(sim, home_team, elo_shift),
      by = c("sim" = "sim", "team" = "home_team")
    ) %>%
    # note that a team on a bye will have `elo_shift` as NA for both joins
    # this means it won't change, which is what we want
    dplyr::mutate(elo = elo + ifelse(!is.na(elo_shift), elo_shift, 0)) %>%
    dplyr::select(-elo_shift)

  # we need to keep `elo_shift` out of `games` too and we're done with it
  games <- games %>%
    dplyr::select(-elo_shift)

  # return the updated teams and games information
  # note that `teams` will now have an updated `elo` column which will
  # be used for the next week's games
  # note that starting `elo` values are the same per-team... 
  # ... but after that will differ per sim depending on that sim's results
  return(list(teams = teams, games = games))
}
```

Let's generate initial random Elo values for each team. To see how this works,
we'll supply an `test_week = 3` as an argument into `simulate_nfl()` which will
abort after simulating Week 3, and instead return the result of our `elo_model()`
function. 

```{r}
initial_elo <- tibble::tibble(
  team = unique(nflseedR::divisions$team),
  elo = rnorm(length(unique(nflseedR::divisions$team)), 1500, 150)
)
test <- simulate_nfl(
  nfl_season = 2020,
  process_games = elo_model,
  elo = initial_elo,
  fresh_season = TRUE, 
  sims_per_round = 1000,
  test_week = 3
)
test$teams %>% dplyr::filter(team == "CHI") %>% utils::head() %>% knitr::kable()
```

You can see the Bears' Elo rating after Week 3 in each of the first six simulations.
The second simulation has been the worst of these six, while the fifth is the highest.

Let's check out the results of those games:

~ RESUMED EDITING HERE ~

## Simulation Configuration

There is a lot of flexibility in how you choose to run the simulation. These are
the parameters and how to configure them when you run the `simulate_nfl()` function.

* **nfl_season** - Which NFL season are you simulating? By default, it simulates
the most recent season for which the regular season schedule is available through
[Lee Sharpe's NFL game data](https://github.com/leesharpe/nfldata/tree/master/data).
This data only goes back to the 1999 season.
* **process_games** - This is where you supply a function you've written to encompass
your model used to determine simulated games results, like the examples above. By
default, this will generate a random Elo for every team per round of simulations, then
use that to determine game data.
* **playoff_seeds** - How many playoff seeds per conference are used? By default, this
is 7 for seasons 2020 and later, and 6 for earlier seasons.
* **if_ended_today** - This should only be used when running in the middle of the
regular season. It will take all completed games as done, but remove the rest of the
regular season games from the schedule, and begin the playoffs as if everything was
locked in based on the regular season data that exists so far.
* **fresh_season** - You'll see this was set to `TRUE` in all of our examples above.
This setting deletes any playoff games and clears out the results for all regular 
season games, so everything is generated fresh. The default is `FALSE` where all games
that have been completed in real life are treated as locked in, and instead remaining
games are simulated.
* **fresh_playoffs** - Similar to `fresh_season`, except instead when set to `TRUE`,
regular season results remain and only playoff games are deleted and then simulated.
The default is `FALSE` in which case playoff games that are completed are accepted
as they occurred,
* **tiebreaker_depth** - How far do you want tiebreakers to be utilized? Usually 
leaving it at the default below (`3`) is fine, but before the season starts, you
may wish to have less tie-breaking in order to 
  * `1`: All teams with the same record have any ties broken randomly.
  * `2`: Instead of evaluating common games if that step is reached, break any ties
randomly. But all earlier tiebreakers are handled correctly.
  * `3`: The default. All tiebreakers are handled through strength of schedule are
processed (or through strength of victory for draft pick order). In the unlikely
event of a further tie, it will be broken randomly.
* **test_week** - This will abort after simulating this week number in the simulator.
`simulate_nfl()` instead will return the output of your `process_games()` function. This
is a useful input for debugging your code, but should be left as `NULL` (the default) for
actual simulations. This also means only the first round will be simulated.
* **simulations** - How many simulations should be run? Defaults to 1000.
* **sims_per_round** - The simulator can break things up into chunks of simulated
seasons, process each chunk on its own (called a round), and then aggregate everything
together at the end. The default value determines the number of locally available 
cores and calculates the number of simulations per round to be equal to half of 
the available cores (various benchmarks showed this results in optimal performance
in parallel processes). If your computer is hanging and forces a restart while 
running this simulation, it is recommended that you lower this number.
