---
title: "Get started with nflseedR"
author: "Sebastian Carl and Lee Sharpe"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.kable.NA = '-')
```

# Preface

nflseedR essentially performs two tasks:

1. Calculation of NFL standings with `nfl_standings()` based on game results of one or more seasons, especially taking into account the comprehensive and sometimes complicated tie-breaking procedures for division ranks, conference seeds and the draft order. Read [this article](https://nflseedr.com/articles/tiebreaker.html) for further information on the implemented tie-breakers.
2. Running thousands of simulations (Monte Carlo style) of an NFL season with `nfl_simulations()`. The standings from point 1 and especially the conference seeds are needed to determine playoff participants. Basically, the first point only exists because we need it to carry out the simulations.

The actual core of a simulation is the generation of game results based on any information that the user deems important. This is why nflseedR is virtually extensible. By default, a simple ELO model is implemented that works with initial starting ELO values and updates them from week to week based on game results. However, the user can write their own function for calculating game results and pass it to nflseedR together with any additional data that may be required.

# Usage

## Standings

We need real or simulated match data to determine standings. The required variables are specified in the documentation of the function `nfl_standings()`.

Here are games data from the 2021 and 2022 seasons.

```{r}
games <- nflreadr::load_schedules(2021:2022)
```

We can pass this data directly to nflseedR and calculate standings. It defaults to compute division ranks as well as conference ranks for all teams and it applies tiebreakers through strength of schedule.

```{r}
standings <- nflseedR::nfl_standings(games)
knitr::kable(standings, digits = 3)
```

## Simulations

With nflseedR 2.0, we have rethought and implemented the execution of simulations from scratch. Particular attention was paid to flexibility and performance. As the usage of the new function `nfl_simulations()` differs from the old function `simulate_nfl()`, we will keep both variants for the time being and maintain two separate articles explaining how to use them.

It is strongly recommended to switch to `nfl_simulations()` because it is far superior to the old implementation in practically every respect, especially in terms of performance.

- Go to [this article](https://nflseedr.com/articles/nflsim2.html) for a detailed explanation of how to use `nfl_simulations()`
- Go to [this article](https://nflseedr.com/articles/nflsim.html) for a detailed explanation of how to use `simulate_nfl()`
